---
title: "Occupancy example walkthrough"
author: "Rob Cooke & Nick Isaac"
date: "12/05/2023"
output: html_notebook
---

## Overview of notebook ##

This notebook will walkthrough preparing data (from <a href = "https://nbnatlas.org/" target = "_blank">NBN Atlas</a>) and building an occupancy model for <a href = "https://www.bwars.com/ant/formicidae/formicinae/lasius-niger" target = "_blank">*Lasius niger*</a> (Small Black Ant) across the UK. Although the model will be built for a single species we will use the target-group approach for the detection model and therefore we will use records for all Formicidae (Ants).

## Set-up ##

Here we load the necessary packages

```{r}

# # install JAGS in terminal
# sudo apt install jags

# # sparta package from github
# remotes::install_github("BiologicalRecordsCentre/sparta")

# load necessary packages
library(sparta)
library(dplyr)
library(R2jags)

# all other packages are called directly, e.g., readr::
# other packages needed: 
# readr
# skimr
# janitor

# global parameters
first_year <- 1970
last_year <- 2023
focal_taxa <- "Lasius niger"

```

## Raw data ##

Download  data for *Formicidae* (Ants) from <a href = "https://nbnatlas.org/" target = "_blank">NBN Atlas</a>

N.B. you need to be logged in to your NBN Atlas account to be able to download data

I excluded records with 'unconfirmed identifications' and 'fossil records'

Load raw downloaded data (downloaded Mon May 15 15:12:40 UTC 2023)

Data is licensed by CC-BY-NC, CC-BY, CC0 or OGL, see the <a href = "https://docs.nbnatlas.org/data-licenses/" target = "_blank">NBN Licences page</a> for more details

```{r}

# original raw records
recs_raw_orig <- readr::read_csv("data/raw/formicidae_records-2023-05-15.csv", col_types = list(`Abundance scale` = readr::col_skip())) # readr performs safer data reading (e.g., dates are read as dates, strings are read as character)
# skip Abundance scale column which is problematic

# overview of data and columns
skimr::skim(recs_raw_orig) %>%
  dplyr::select(skim_type, skim_variable, n_missing)

# tidy data
recs <- recs_raw_orig %>% 
  # tidy column names
  janitor::clean_names() %>% 
  # only species-level records
  dplyr::filter(taxon_rank == "species") %>% 
  # only records recorded to day precision
  dplyr::filter(!is.na(start_date)) %>%
  dplyr::filter(is.na(end_date)) %>% 
  # only records recorded to 4-figure grid ref (1km)
  dplyr::filter(!is.na(osgr_1km)) %>% 
  # get year from date
  dplyr::mutate(year = as.numeric(format(start_date, "%Y"))) %>% 
  # temporal window
  dplyr::filter(year >= first_year & year <= last_year) %>% 
  # remove duplicates
  dplyr::distinct(scientific_name, start_date, osgr_1km, year)

# number of records after cleaning
nrow(recs)

# number of records for Lasius niger
nrow(dplyr::filter(recs, scientific_name == focal_taxa))

```

## Data diagnostics ##

Run some data diagnostics within sparta

The plot produced shows the number of records for each year in the top plot and the average list length in a box plot at the bottom. List length is the number of taxa observed on a visit to a site, where a visit is taken to be a unique combination of ‘where’ and ‘when’. A trend in the number of observations across time is not uncommon and a formal test for such a trend is performed in the form of a linear model. Trends in the number of records over time are handled by all of the methods presented in sparta in a variety of different ways. A trend in list length can cause some methods such as the reporting rate methods to fail (see ‘LessEffortPerVisit’ scenario in <a href = "https://doi.org/10.1111/2041-210X.12254" target = "_blank">Isaac et al. (2014)</a>

```{r}

# run data diagnostics - time period is year
dd <- sparta::dataDiagnostics(taxa = recs$scientific_name,
                              site = recs$osgr_1km,
                              time_period = recs$year,
                              progress_bar = FALSE)

```

## Format the data for sparta ##

Here we format the data to match the structure needed to run a 'sparta' occupancy model

```{r}

vis_data <- sparta::formatOccData(taxa = recs$scientific_name,
                                  site = recs$osgr_1km,
                                  survey = recs$start_date)

# saveRDS(vis_data, "data/processed/vis_data_Ants.rds")

skimr::skim(vis_data) %>% 
  # remove unnecessary columns
  dplyr::select(-n_missing, -complete_rate, -character.min, -character.max, -character.empty, -character.n_unique, -character.whitespace)

```

## Run sparta occupancy model ##

Here we run a 'sparta' occupancy model

Specifically, we fit a multi-season Bayesian occupancy model (Outhwaite et al., 2018, Outhwaite et al., 2019) to the occurrence records for each species. These models separate occupancy (the proportion of occupied 1 km grid cells) and detection into hierarchically coupled submodels to allow for imperfect detection and temporal changes in recorder intensity, which are common biases in occurrence record datasets (Isaac and Pocock, 2015). 

We used a closure period (the temporal precision of the occupancy submodel) of one year (Outhwaite et al., 2018). We therefore estimated occupancy annually between `r first_year` and `r last_year`, using a separate model for each species. 

The detection submodel estimates the probability of detection based on repeat visits (a visit is a unique combination of 1 km grid cell and day) within years. As the data are presence-only, we used records of other species within the same taxonomic group (the target-group approach) to infer non-detections of the focal species (Kéry et al., 2010). 

We included the number of species recorded during a visit, categorized into lists of 1, 2–3 or 4+ records, in the detection submodel to estimate sampling intensity and variability in selective reporting (Outhwaite et al., 2019; Szabo et al., 2010; van Strien et al., 2013). 

We also added a random effect (intercept) of grid cell to allow for variation in occupancy status and uneven sampling among grid cells (Isaac et al., 2014; Kuussaari et al., 2007).

For the priors, we selected a random walk prior on the year effect, which enabled the sharing of information between the current and previous year in the occupancy submodel - imposing an *a priori* judgement that a species' occupancy is likely to be similar from one year to the next (Outhwaite et al., 2018, Outhwaite et al., 2019). 

We used uninformative priors for the remaining parameters within the model, following Outhwaite et al. (2019).

We fit occupancy models using the occDetFunc function in the sparta package (August et al., 2020), which uses a Markov Chain Monte Carlo algorithm to fit the models via JAGS (Plummer, 2013). 

We specified three chains, 32,000 iterations, a burn in of 30,000, and a thinning rate of six. 

```{r}

mod_out <- sparta::occDetFunc(taxa_name = as.character(focal_taxa),
                              occDetdata = vis_data$occDetdata,
                              spp_vis = vis_data$spp_vis,
                              write_results = TRUE,
                              output_dir = "data/processed/",
                              n_chains = 3,
                              n_iterations = 32000,
                              burnin = 30000,
                              thinning = 6,
                              nyr = 2,
                              modeltype = c('ranwalk', 'halfcauchy', 'catlistlength'),
                              provenance = "NBN Atlas Formicidae Mon May 15 15:12:40 UTC 2023. vis_data_Ants.rds",
                              return_data = FALSE,
                              seed = 842845)

```

## Checking the model ##

We determined convergence based on the Gelman-Rubin statistic (Rhat < 1.1)

```{r}



```

## Summarizing the results ##

```{r}



```


## Resources ##

**Useful resources for sparta-style occupancy models:**<br/>
Isaac, N.J., van Strien, A.J., August, T.A., de Zeeuw, M.P. and Roy, D.B., 2014.
<a href = "https://doi.org/10.1111/2041-210X.12254" target = "_blank">Statistics for citizen science: extracting signals of change from noisy ecological data</a>, pp.1052-1060.<br/>

Outhwaite, C.L., Chandler, R.E., Powney, G.D., Collen, B., Gregory, R.D. and Isaac, N.J., 2018. <a href = "https://doi.org/10.1016/j.ecolind.2018.05.010" target = "_blank">Prior specification in Bayesian occupancy modelling improves analysis of species occurrence data</a>. Ecological Indicators, 93, pp.333-343.<br/>

Outhwaite, C.L., Powney, G.D., August, T.A., Chandler, R.E., Rorke, S., Pescott, O.L., Harvey, M., Roy, H.E., Fox, R., Roy, D.B. and Alexander, K., 2019. <a href = "https://www.nature.com/articles/s41597-019-0269-1" target = "_blank">Annual estimates of occupancy for bryophytes, lichens and invertebrates in the UK, 1970–2015</a>. Scientific data, 6(1), p.259.<br/>

Outhwaite, C.L., Gregory, R.D., Chandler, R.E., Collen, B. and Isaac, N.J., 2020. <a href = "https://www.nature.com/articles/s41559-020-1111-z" target = "_blank">Complex long-term biodiversity change among invertebrates, bryophytes and lichens</a>. Nature ecology & evolution, 4(3), pp.384-392.<br/>

Cooke, R., Mancini, F., Boyd, R.J., Evans, K.L., Shaw, A., Webb, T.J. and Isaac, N.J., 2023. <a href = "https://doi.org/10.1016/j.biocon.2022.109884" target = "_blank">Protected areas support more species than unprotected areas in Great Britain, but lose them equally rapidly</a>. Biological Conservation, 278, p.109884.<br/>

Boyd, R.J., August, T.A., Cooke, R., Logie, M., Mancini, F., Powney, G.D., Roy, D.B., Turvey, K. and Isaac, N.J., 2023. <a href = "https://onlinelibrary.wiley.com/doi/full/10.1111/brv.12961" target = "_blank">An operational workflow for producing periodic estimates of species occupancy at national scales</a>. Biological Reviews.
